---
title: "ST606 Data Visualizations"
author: 'Dennis Thomas Veliyath'
date: "12/04/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- packages used -->

```{r packages,warning=FALSE,message=FALSE}
library(stringr)
library(tidyr)
library(forcats)
library(dplyr)
library(countrycode)
library(ggplot2)
library(showtext)
library(tidyverse)
library(gganimate)
```

## Loading the data

```{r loadingdata,warning=FALSE,message=FALSE}

#Loading birth data and storing in a temp variable

birth_data <- readr::read_csv('~/VSA01.20220412164432.csv')
temp_birth_data<-birth_data

#Loading boy names data and storing in a temp variable

boys_names_data<-readr::read_csv('~/VSA50.20220412164235.csv')
temp_boys_data<-boys_names_data

#Loading girl names data and storing in a temp variable

girls_names_data<-readr::read_csv('~/VSA60.20220412164418.csv')
temp_girls_data<-girls_names_data

font_add_google("Quattrocento Sans", "Quattrocento Sans")
showtext_auto()
windows()

```

### 1. Plotting to display patterns of a particular name over the years.


```{r}

# Plot to display trend of Dennis over the years based on occurences

graph_data <- temp_boys_data %>% filter( Names== "Dennis", STATISTIC == "VSA50C01") %>% drop_na(VALUE) 

gg_count_title <- expression(paste("The name ", italic(bold("Dennis")), " over the years based on number of occurences"))

ggplot(graph_data, aes(x=Year, y=VALUE))+  
  geom_bar(stat="identity",fill = "darkblue") +
  theme_classic() + 
  labs(title = gg_count_title, x = "Year", y = "Count")+
  scale_x_continuous(breaks = graph_data$Year, labels = graph_data$Year)+ 
  scale_y_continuous(breaks = graph_data$VALUE, labels = graph_data$VALUE) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1,size = 22),axis.text.y = element_text(size = 22),axis.title.x = element_text(size = 22),axis.title.y = element_text(size = 22), plot.title = element_text(family="Quattrocento Sans", hjust = 0.5,size = 30))

```


```{r}

# Plot to display trend of Dennis over the years based on rank

graph_rank_data <- temp_boys_data %>% filter( Names== "Dennis", STATISTIC == "VSA50C02") %>% drop_na(VALUE) 

gg_rank_title <- expression(paste("The name ", italic(bold("Dennis")), " over the years based on rank"))

ggplot(graph_rank_data, aes(x=Year, y=VALUE))+  
  geom_bar(stat="identity",fill = "darkblue") +
  theme_classic() + 
  labs(title = gg_rank_title, x = "Years", y = "Rank")+
  scale_x_continuous(breaks = graph_rank_data$Year, labels = graph_rank_data$Year)+ 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1,size = 22),axis.text.y = element_text(size = 22),axis.title.x = element_text(size = 22),axis.title.y = element_text(size = 22), plot.title = element_text(family="Quattrocento Sans", hjust = 0.5,size = 30))


```

### 2: Top 10 names over the years

```{r,warning=FALSE}

#Top Boy names over the years
count_data <- temp_boys_data %>% filter(STATISTIC == "VSA50C01",Year=="1964") %>% drop_na(VALUE)
count_test_data <- temp_boys_data %>% filter(STATISTIC == "VSA50C01") %>% drop_na(VALUE)
sort_data<-count_data[with(count_data,order(-VALUE)),]
sort_data <- sort_data[1:10,]


name_formatted <- count_test_data %>%
  group_by(Year) %>%
  mutate(rank = rank(-VALUE),
         Value_rel = VALUE/VALUE[rank==1],
         Value_lbl = paste0(" ",round(VALUE/1e9))) %>%
  group_by(Names) %>%
  filter(rank <=10) %>%
  ungroup()


staticplot = ggplot(name_formatted, aes(rank, group = Names,
                fill = as.factor(Names), color = as.factor(Names))) +
  geom_tile(aes(y = VALUE/2,
                height = VALUE,
                width = 0.9), alpha = 0.8, color = NA) +
  geom_text(aes(y = 0, label = paste(Names, " ")), vjust = 0.2, hjust = 1) +
  geom_text(aes(y=VALUE,label = VALUE, hjust=0)) +
  coord_flip(clip = "off", expand = FALSE) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_reverse() +
  guides(color = FALSE, fill = FALSE) +
  theme(axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        legend.position="none",
        panel.background=element_blank(),
        panel.border=element_blank(),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.grid.major.x = element_line( size=.1, color="grey" ),
        panel.grid.minor.x = element_line( size=.1, color="grey" ),
        plot.title=element_text(size=25, hjust=0.5, face="bold", colour="grey", vjust=1),
        plot.subtitle=element_text(size=18, hjust=0.5, face="italic", color="grey"),
        plot.caption =element_text(size=10, hjust=0.5, face="italic", color="grey", vjust = 1),
        plot.background=element_blank(),
        plot.margin = ggplot2::margin(4,4, 4, 8, "cm"))

anim = staticplot + transition_states(Year, transition_length = 8, state_length = 1) +
  view_follow(fixed_x = TRUE)  +
  labs(title = 'Top Boy Names of the Year : {closest_state}\n',
       caption  = "\nMost common names over the years")

animate(anim, nframes = 600, renderer = gifski_renderer("gganim.gif"), width = 1200, height = 1000, fps = 15)

```

```{r,warning=FALSE}

#Top Girl names over the years

count_girlstest_data <- temp_girls_data %>% filter(STATISTIC == "VSA60C01") %>% drop_na(VALUE)
sort_data<-count_data[with(count_data,order(-VALUE)),]
sort_data <- sort_data[1:10,]


name_girl_formatted <- count_girlstest_data %>%
  group_by(Year) %>%
  mutate(rank = rank(-VALUE),
         Value_rel = VALUE/VALUE[rank==1],
         Value_lbl = paste0(" ",round(VALUE/1e9))) %>%
  group_by(Names) %>%
  filter(rank <=10) %>%
  ungroup()


staticgirlplot = ggplot(name_girl_formatted, aes(rank, group = Names,
                fill = as.factor(Names), color = as.factor(Names))) +
  geom_tile(aes(y = VALUE/2,
                height = VALUE,
                width = 0.9), alpha = 0.8, color = NA) +
  geom_text(aes(y = 0, label = paste(Names, " ")), vjust = 0.2, hjust = 1) +
  geom_text(aes(y=VALUE,label = VALUE, hjust=0)) +
  coord_flip(clip = "off", expand = FALSE) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_reverse() +
  guides(color = FALSE, fill = FALSE) +
  theme(axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        legend.position="none",
        panel.background=element_blank(),
        panel.border=element_blank(),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.grid.major.x = element_line( size=.1, color="grey" ),
        panel.grid.minor.x = element_line( size=.1, color="grey" ),
        plot.title=element_text(size=25, hjust=0.5, face="bold", colour="grey", vjust=1),
        plot.subtitle=element_text(size=18, hjust=0.5, face="italic", color="grey"),
        plot.caption =element_text(size=10, hjust=0.5, face="italic", color="grey", vjust = 1),
        plot.background=element_blank(),
        plot.margin = ggplot2::margin(4,4, 4, 8, "cm"))

animgirl = staticgirlplot + transition_states(Year, transition_length = 8, state_length = 1) +
  view_follow(fixed_x = TRUE)  +
  labs(title = 'Top Girl Names of the Year : {closest_state}\n',
       caption  = "\nMost common girl names over the years")

animate(animgirl, nframes = 600, renderer = gifski_renderer("gganimgirl.gif"), width = 1200, height = 1000, fps = 15)

```


### 3: Registered births over the years

```{r}

# No of total births registered over the years

birth_temp_data <- temp_birth_data %>% filter(Sex=="Both sexes") %>% drop_na(VALUE) 
birth_male_temp_data <- temp_birth_data %>% filter(Sex=="Male") %>% drop_na(VALUE) 
birth_female_temp_data <- temp_birth_data %>% filter(Sex=="Female") %>% drop_na(VALUE) 

gg_birth_title <- expression(paste("The number of total births over the years (1910-2020)"))
gg_male_birth_title <- expression(paste("The number of male births over the years (1910-2020)"))
gg_female_birth_title <- expression(paste("The number of female births over the years (1910-2020)"))

ggplot(birth_temp_data, aes(x=Year, y=VALUE))+  
  geom_bar(stat="identity",fill = "#000000") +
  theme_classic() + 
  labs(title = gg_birth_title, x = "Years", y = "Count")+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1,size = 18),axis.text.y = element_text(size = 22),axis.title.x = element_text(size = 22),axis.title.y = element_text(size = 22), plot.title = element_text(family="Quattrocento Sans", hjust = 0.5,size = 30))

```

```{r}
min(birth_temp_data[,8])
min_val_year <- birth_temp_data[birth_temp_data$VALUE == 48255, "Year"]
```


Lowest births recorded was in 1994 with only 48255 births(both sexes). This was due to the 1980-1987 long recession which lead to poor living conditions throughout the mid 1990's. The recession also cause a majority of people to emigrate from Ireland in search of a better life.




## Male births over the years

```{r}

# No of total male births registered over the years

ggplot(birth_male_temp_data, aes(x=Year, y=VALUE))+  
  geom_bar(stat="identity",fill = "darkblue") +
  theme_classic() + 
  labs(title = gg_male_birth_title, x = "Years", y = "Count")+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1,size = 18),axis.text.y = element_text(size = 22),axis.title.x = element_text(size = 22),axis.title.y = element_text(size = 22), plot.title = element_text(family="Quattrocento Sans", hjust = 0.5,size = 30))

```

## Female Births over the years

```{r}

# No of total female births registered over the years

ggplot(birth_female_temp_data, aes(x=Year, y=VALUE))+  
  geom_bar(stat="identity",fill = "deeppink") +
  theme_classic() + 
  labs(title = gg_female_birth_title, x = "Years", y = "Count")+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1,size = 18),axis.text.y = element_text(size = 22),axis.title.x = element_text(size = 22),axis.title.y = element_text(size = 22), plot.title = element_text(family="Quattrocento Sans", hjust = 0.5,size = 30))

```


### 4: Total names(count>3) over the years

```{r}

# Total boy's names vs years

count_boys_test_data<- temp_boys_data %>% filter(STATISTIC == "VSA50C01") %>% drop_na(VALUE)

ggplot(count_boys_test_data, aes(x=Year))+  
  geom_bar(stat="count",fill = "darkblue") +
  theme_classic() + 
  labs(title = "Total Boy's Names vs Years", x = "Years", y = "Count")+
  scale_x_continuous(breaks = scales::breaks_width(4))+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1,size = 22),axis.text.y = element_text(size = 22),axis.title.x = element_text(size = 22),axis.title.y = element_text(size = 22), plot.title = element_text(family="Quattrocento Sans", hjust = 0.5,size = 30))

```



```{r}

# Total girl's names vs years

count_girls_test_data <- temp_girls_data %>% filter(STATISTIC == "VSA60C01") %>% drop_na(VALUE) 
max(count_girls_test_data$VALUE)
ggplot(count_girls_test_data, aes(x=Year))+  
  geom_bar(stat="count",fill = "deeppink") +
  theme_classic() + 
  labs(title = "Total Girl's Names vs Years", x = "Years", y = "Count")+
  scale_x_continuous(breaks = scales::breaks_width(4))+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1,size = 22),axis.text.y = element_text(size = 22),axis.title.x = element_text(size = 22),axis.title.y = element_text(size = 22), plot.title = element_text(family="Quattrocento Sans", hjust = 0.5,size = 30))
```

## Total boys and girls named vs Total boys and girls registered births over the years


```{r}

birth_male_temp_data_new<- birth_male_temp_data[birth_male_temp_data$Year >= '1964',]

birth_female_temp_data_new<- birth_female_temp_data[birth_female_temp_data$Year >= '1964',]

df_boys<- count_boys_test_data %>% group_by(Year) %>%
  summarise(VALUE = sum(VALUE))

df_girls<- count_girls_test_data %>% group_by(Year) %>%
  summarise(VALUE = sum(VALUE))

combined_plot<- ggplot() +
      # orange plot
      geom_point(data=birth_male_temp_data_new, aes(x=Year, y=VALUE, color="boy_births", shape = "boy_births" ),size=1.5) +
      # green plot
      geom_point(data=birth_female_temp_data_new, aes(x=Year, y=VALUE, color="girl_births", shape = "girl_births"), size=1.5) +
      # blue plot
      geom_point(data=df_boys, aes(x=Year, y=VALUE, color="boys_named", shape= "boys_named"), size=1.5) +
      #pink plot
      geom_point(data=df_girls, aes(x=Year, y=VALUE, color="girls_named", shape="girls_named"), size=1.5) + 
  
      scale_x_continuous(breaks = scales::breaks_width(4)) +
  
      scale_y_continuous(breaks = scales::breaks_width(2500)) +
  
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=0.5,size = 12))+
  
      scale_colour_manual(name="legend", values = c("boy_births"= "darkblue", "girl_births"="green","boys_named"="orange", "girls_named"="deeppink")) +
  
      scale_shape_manual(name="legend", values = c("boy_births" = 15, "girl_births" = 16,"boys_named"=15, "girls_named"=16 ))

combined_plot

```

## Most Used Letter for boys over the Years


```{r}


# Most used letter for boys over the years

letter_data <- temp_boys_data %>% filter(STATISTIC == "VSA50C01")%>% drop_na(VALUE)

# Adding new column and assigning values
letter_data <- letter_data %>% add_column(new_col = NA)
names(letter_data)[9]<- 'Letter_group'

letter_data$Letter_group[startsWith(letter_data$Names,"A")] <- "A"
letter_data$Letter_group[startsWith(letter_data$Names,"B")] <- "B"
letter_data$Letter_group[startsWith(letter_data$Names,"C")] <- "C"
letter_data$Letter_group[startsWith(letter_data$Names,"D")] <- "D"
letter_data$Letter_group[startsWith(letter_data$Names,"E")] <- "E"
letter_data$Letter_group[startsWith(letter_data$Names,"F")] <- "F"
letter_data$Letter_group[startsWith(letter_data$Names,"G")] <- "G"
letter_data$Letter_group[startsWith(letter_data$Names,"H")] <- "H"
letter_data$Letter_group[startsWith(letter_data$Names,"I")] <- "I"
letter_data$Letter_group[startsWith(letter_data$Names,"J")] <- "J"
letter_data$Letter_group[startsWith(letter_data$Names,"K")] <- "K"
letter_data$Letter_group[startsWith(letter_data$Names,"L")] <- "L"
letter_data$Letter_group[startsWith(letter_data$Names,"M")] <- "M"
letter_data$Letter_group[startsWith(letter_data$Names,"N")] <- "N"
letter_data$Letter_group[startsWith(letter_data$Names,"O")] <- "O"
letter_data$Letter_group[startsWith(letter_data$Names,"P")] <- "P"
letter_data$Letter_group[startsWith(letter_data$Names,"Q")] <- "Q"
letter_data$Letter_group[startsWith(letter_data$Names,"R")] <- "R"
letter_data$Letter_group[startsWith(letter_data$Names,"S")] <- "S"
letter_data$Letter_group[startsWith(letter_data$Names,"T")] <- "T"
letter_data$Letter_group[startsWith(letter_data$Names,"U")] <- "U"
letter_data$Letter_group[startsWith(letter_data$Names,"V")] <- "V"
letter_data$Letter_group[startsWith(letter_data$Names,"W")] <- "W"
letter_data$Letter_group[startsWith(letter_data$Names,"X")] <- "X"
letter_data$Letter_group[startsWith(letter_data$Names,"Y")] <- "Y"
letter_data$Letter_group[startsWith(letter_data$Names,"Z")] <- "Z"

letter_data<- letter_data%>% drop_na(Letter_group)

ggplot(letter_data, aes(x=letter_data$Letter_group))+  
  geom_bar(stat="count",fill = "darkblue") +
  theme_classic() + 
  labs(title = "Most Used Letters For Boys Over The Years", x = "Letters", y = "Count")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1,size = 22),axis.text.y = element_text(size = 22),axis.title.x = element_text(size = 22),axis.title.y = element_text(size = 22), plot.title = element_text(family="Quattrocento Sans", hjust = 0.5,size = 30))


```

Here we can see that D is the most used initial for first names over the years and X is the least used initial.

## Trend for boy names starting with 'D' over the years 

```{r}

# Trend for boy names starting with 'D' over the years 

starts_with_d <- temp_boys_data %>% filter(STATISTIC == "VSA50C01")%>% drop_na(VALUE)
names_with_d<-starts_with_d[grepl("^D", starts_with_d$Names), ]
names_with_d<- names_with_d[,-c(1,2,3,5,7)]

ggplot(names_with_d, aes(x=Year))+  
  geom_point(stat="count",colour = "darkblue") +
  theme_classic() + 
  labs(title = "Trend for boy  names starting with 'D' over the years", x = "Year", y = "Count")+
  scale_x_continuous(breaks = scales::breaks_width(4))+
  
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1,size = 22),axis.text.y = element_text(size = 22),axis.title.x = element_text(size = 22),axis.title.y = element_text(size = 22), plot.title = element_text(family="Quattrocento Sans", hjust = 0.5,size = 30))

```


## Most used letter for girls over the years

```{r}
# Most Used Letter for girls over the Years

letter_data_girls <- temp_girls_data %>% filter(STATISTIC == "VSA60C01")%>% drop_na(VALUE)

# Adding new column and assigning values

letter_data_girls <- letter_data_girls %>% add_column(new_col = NA)
names(letter_data_girls)[9]<- 'Letter_group'

letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"A")] <- "A"
letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"B")] <- "B"
letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"C")] <- "C"
letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"D")] <- "D"
letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"E")] <- "E"
letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"F")] <- "F"
letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"G")] <- "G"
letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"H")] <- "H"
letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"I")] <- "I"
letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"J")] <- "J"
letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"K")] <- "K"
letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"L")] <- "L"
letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"M")] <- "M"
letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"N")] <- "N"
letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"O")] <- "O"
letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"P")] <- "P"
letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"Q")] <- "Q"
letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"R")] <- "R"
letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"S")] <- "S"
letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"T")] <- "T"
letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"U")] <- "U"
letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"V")] <- "V"
letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"W")] <- "W"
letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"X")] <- "X"
letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"Y")] <- "Y"
letter_data_girls$Letter_group[startsWith(letter_data_girls$Names,"Z")] <- "Z"

letter_data_girls<- letter_data_girls%>% drop_na(Letter_group)

ggplot(letter_data_girls, aes(x=letter_data_girls$Letter_group))+  
  geom_bar(stat="count",fill = "deeppink") +
  theme_classic() + 
  labs(title = "Most Used Letters for Girls Over The Years", x = "Letters", y = "Count")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1,size = 22),axis.text.y = element_text(size = 22),axis.title.x = element_text(size = 22),axis.title.y = element_text(size = 22), plot.title = element_text(family="Quattrocento Sans", hjust = 0.5,size = 30))

```


## Trend for girl names starting with 'A' over the years 

```{r}

# Trend for girl names starting with 'A' over the years 

starts_with_a <- temp_girls_data %>% filter(STATISTIC == "VSA60C01")%>% drop_na(VALUE)
names_with_a<-starts_with_a[grepl("^A", starts_with_a$Names), ]
names_with_a<- names_with_a[,-c(1,2,3,5,7)]

ggplot(names_with_a, aes(x=Year))+  
  geom_point(stat="count",colour = "deeppink") +
  theme_classic() + 
  labs(title = "Trend for girl names starting with 'A' over the years", x = "Year", y = "Count")+
  scale_x_continuous(breaks = scales::breaks_width(4))+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1,size = 22),axis.text.y = element_text(size = 22),axis.title.x = element_text(size = 22),axis.title.y = element_text(size = 22), plot.title = element_text(family="Quattrocento Sans", hjust = 0.5,size = 30))



```


